---
title: "A1 New: Exploration and Cleaning of Dataset"
author: "Michelle Lin"
date: "12 March 2024"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
bibliography: A1_Bibliography.bib
---

```{r hide=TRUE, message = FALSE, warning = FALSE, include=FALSE}
# Set up and Install all Libraries if not present

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("GEOquery", quietly = TRUE))
    install.packages("GEOquery")

if (!requireNamespace("knitr", quietly = TRUE))
    install.packages("knitr")

if (!requireNamespace("edgeR", quietly = TRUE))
    install.packages("edgeR")

if (!requireNamespace("hrbrthemes", quietly = TRUE))
    install.packages("hrbrthemes")

if (!requireNamespace("ggplot2", quietly = TRUE))
    install.packages("ggplot2")

if (!requireNamespace("tidyr", quietly = TRUE))
    install.packages("tidyr")


```

## IMPORT DATA AND INITIAL EXPLORATION

### Set Global variables and constants
```{r message = FALSE, warning = FALSE}
# Set GEOID 
GEOID <- "GSE160230"

## File location
workingDir <- file.path(getwd())
```



### Study Background Information
The study chosen for this project is "hPSC-derived Airway Organoids-based Screen Reveals the Role of HIF1/ Glycolysis Axis in SARS-CoV-2 Infection" published in 2021 by Duan et al.[@duan2021airway]

The authors conducted a study in human pluripotent stem cell  (hPSC) derived airway organoids - (hPSC-AO) where they studied the molecular mechanism behind SARS-Cov-2 infections and the differential expression of different components of the pathways when treated with GW6571.

The following is a summary of the study chosen for this assignment. 
```{r warning=FALSE, message=FALSE, }
# get information for study

gse <- GEOquery::getGEO(GEOID, GSEMatrix = FALSE)
currentGPL <- names(GEOquery::GPLList(gse))[1]
currentGPLInfo <- GEOquery::Meta(GEOquery::getGEO(currentGPL))

```
**GEO ID of data set:** GSE160230 

**Organism: ** `r currentGPLInfo$organism` 

**Submission Date:** `r gse@header$submission_date`

**Last Updated: ** `r gse@header$last_update_date`

**Brief Summary of data set:** `r gse@header$summary` 

**Platform:** `r currentGPLInfo$title` 

**Platform Last Updated: ** `r currentGPLInfo$last_update_date`


### Get Expression Data Files
```{r message = FALSE, warning = FALSE}
# Set Up + Download Data if not present

filesSupp <- GEOquery::getGEOSuppFiles(GEOID, fetch_files= FALSE)

fileMissing <- filesSupp$fname[!unlist(
  lapply(filesSupp$fname, FUN=function(x){
    file.exists(
      file.path(workingDir,GEOID,x))}))]

if(length(fileMissing) > 0){
  for(i in seq_along(fileMissing)){
    # get the supplementary files
    sfiles = GEOquery::getGEOSuppFiles(GEOID, 
                                       baseDir = workingDir,
                                       fetch_files = TRUE)
  }
}
```

The supplementary files found in this dataset are: 

`r filesSupp$fname` 

For this study, we will pay particular attention to the "GSE160230_raw_counts.hPSC_AO.downsample.txt" i*n particular. 

There were three conditions observed in this study - **3 mock samples** - untreated and uninfected, **3 control samples** - untreated and infected and **3 GW_treated samples** - treated and infected samples. Due to the difference in higher observed viral counts and hence, lower human gene counts in SARS-Cov2 infected samples, the raw counts in the uninfected samples (mock and treated) was down-sampled so that the median human gene count was comparable to that in the infected samples. [@duan2021airway]


## 1. Load and Preprocess Data
```{r message = FALSE, warning = FALSE }
# Read data files
dataFile <- "GSE160230_raw_counts.hPSC_AO.downsample.txt.gz"
  
aoCtrlTreatMat <- utils::read.table(
  file.path(workingDir, GEOID, dataFile),
  header = TRUE,
  check.names = TRUE
)
```

The dataset is of shape (`r dim(aoCtrlTreatMat)`) with 48815 genes and 9 samples- 3 mock (m), 3 untreated and infected (d), 3 treated and infected (gw).
A preview of the dataset can be seen below: 
```{r message = FALSE, warning = FALSE }
knitr::kable(aoCtrlTreatMat[1:7, 1:9], format = "html")
```
```{r message = FALSE, warning = FALSE,, hide=TRUE}
# Initial Pre-processing of data
any(is.na(aoCtrlTreatMat))
```
There were no missing values found in this dataset.

## HUGO GENE SYMBOL MATCH

The following is a table of some of the Ensembl human datasets.
```{r message = FALSE, warning = FALSE}
library(biomaRt)

# Ensembl
ensembl <- biomaRt::useMart("ensembl")
datasetEnsembl <- biomaRt::listDatasets(ensembl)
knitr::kable(head(datasetEnsembl[grep(datasetEnsembl$dataset, pattern = "sapiens"), ]), format = "html")
ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", mart=ensembl)
```

```{r}
# filters
biomartHumanFilters <- biomaRt::listFilters(ensembl) 
knitr::kable(biomartHumanFilters[
  grep(biomartHumanFilters$name, pattern="ensembl"), ], format = "html")

# create conversion object if not present and save
stashConversion <- "ensemblHUGOConversion.rds"
ids2convert <- rownames(aoCtrlTreatMat)
if(file.exists(stashConversion)){
  conversionLink <- readRDS(stashConversion)
} else {
  conversionLink <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                          filters = c("ensembl_gene_id"),
                          values = ids2convert,
                          mart =ensembl )
  saveRDS(conversionLink, stashConversion)
}


# Merge identifier with original table
objAnnotated <- merge(conversionLink, aoCtrlTreatMat, by.x = 1, by.y = 0, all.y = TRUE)

```



### Remove Unmapped Genes
```{r message = FALSE, warning = FALSE}
# Find the number of genes that remained unmatched

## number of genes mapped
mappedGenes <- length(which(rownames(aoCtrlTreatMat) %in% 
               conversionLink$ensembl_gene_id))

## total num_genes 
totalGenes <- nrow(aoCtrlTreatMat)

## number of genes missing
unmatchedGenes <- totalGenes - mappedGenes

```
There were `r mappedGenes` mapped human genes of the total `r totalGenes` transcribed and `r unmatchedGenes` genes remained unmatched. This is potentially due to 2 main reasons - outdated Ensembl mappings and the remaining genes being viral genes which we are not interested in. 

```{r message = FALSE, warning = FALSE}
## show table
unmatched <- objAnnotated[which(is.na(objAnnotated$hgnc_symbol)), ]
unmatchedEg <- knitr::kable(objAnnotated[which(is.na(objAnnotated$hgnc_symbol))[1:5], 1:5], type = "html")

matchObj <- objAnnotated[!(is.na(objAnnotated$hgnc_symbol)), ]
```
There were `r length(unmatched$ensembl_gene_id)` genes that remained unmatched. Some examples of which are: `r unmatchedEg`.


### Combine Rows with the same Gene Names
```{r message = FALSE, warning = FALSE}
# create a table for HUGO genes on row and sample on column
uniqGenes <- length(unique(matchObj$hgnc_symbol))
uniqEnblIDs <- length(unique(matchObj$ensembl_gene_id))
```
There were found to be `r uniqGenes` unique gene mappings and `r uniqEnblIDs` ensembl gene IDs mapped, indicating that several of the Ensembl gene IDs mapped to the same gene. Hence, next, we will be aggregating these rows. 

```{r message = FALSE, warning = FALSE}
# Aggregate duplicated genes

## find duplicated genes and unique genes
duplicatedGenes <- unique(matchObj$hgnc_symbol[duplicated(matchObj$hgnc_symbol)])
uniqueGenes <- unique(matchObj$hgnc_symbol)

## aggregation and remove un-named gene
matchObj <- matchObj[, -which(colnames(matchObj) == "ensembl_gene_id")]
aggmatchObj <- aggregate(.~hgnc_symbol, data = matchObj, FUN=sum)
aggmatchObj <- aggmatchObj[-c(aggmatchObj$hgnc_symbol == ""), ]

## save intermediary objects
saveRDS(aggmatchObj, "AggregatedDataset.RDS")
write.csv(aggmatchObj, "AggregatedDataset.csv")
```

Our final dataset after cleaning the data results in a dataset of shape (`r dim(aggmatchObj)`), that is 33833 labelled genes and expressions values for 9 samples - 3 mock, 3 untreated and infected, 3 treated and infected.



## DATA CLEANING AND PRE-PROCESSING

### Remove Low Expression Genes
```{r message = FALSE, warning = FALSE}
# choose min number of samples
min_samples <- 3

aggMat <- as.matrix(aggmatchObj)
rownames(aggMat) <- aggMat[, 1]
aggMat <- aggMat[,-1]

aggMat <- trimws(aggMat)
write.csv(aggMat, "aggregated_matrix.csv")

for (ri in seq_along(rownames(aggMat))){
  for (cj in seq_along(colnames(aggMat))){
    aggMat[ri, cj] <- as.integer(aggMat[ri, cj])
  }
}

intMat <- matrix(as.integer(aggMat), nrow = nrow(aggMat), ncol = ncol(aggMat))
colnames(intMat) <- c("Mock1", "Mock2", "Mock3", "Control1", "Control2", "Control3", "GW-Treated1", "GW-Treated2", "GW-Treated3")
rownames(intMat) <- rownames(aggMat)

satisfied <- rowSums(edgeR::cpm(int_matrix) > 1) >= min_samples
cleanData <- intMat[satisfied, ]

write.csv(cleanData, "CleanDataset.csv")
saveRDS(cleanData, "CleanDataset.RDS")
```
From the above, we removed `r (33833 - sum(satisfied))`  genes due to low expression values.


```{r message = FALSE, warning = FALSE}
# Plot the data before and after
library(hrbrthemes)
library(ggplot2)


## BOXPLOT
## Add 1 to each value in the cleanData to prevent -inf errors for 0 expression
dataAdd1 <- cleanData + 1
dataLog <- log2(edgeR::cpm(dataAdd1))

dfLog <- as.data.frame(dataLog)
dfLog$Gene <- rownames(dfLog)
dfLogLonger <- tidyr::pivot_longer(dfLog, cols = -Gene, names_to = "Sample", values_to = "Log2CPM")

ggplot2::ggplot(dfLogLonger, aes(x = Sample, y = Log2CPM)) +
  geom_boxplot() + 
  labs(x = "Samples", y = "Log2 CPM") +
  ggtitle("Boxplot showing Log2 CPM Prenormalisation")
  
## MDS PLOT
limma::plotMDS(log2(cleanData), labels=NULL, pch=1,
               col = c("red", "green", "blue")[factor(groups)])

legend("topright",
  legend=levels(factor(groups)),
  pch=c(1), col= c("red", "green","blue"),title="Class",
  bty = 'n', cex = 0.75)

## DENSITY PLOT
## Add 0.1 to each value in the cleanData to prevent 0 expression error
dataAddp1 <- cleanData + 0.1
dataLogp1 <- log2(edgeR::cpm(dataAddp1))

dfLogp1 <- as.data.frame(dataLogp1)
dfLogp1$Gene <- rownames(dfLogp1)
dfLogLongerp1 <- tidyr::pivot_longer(dfLogp1, cols = -Gene, names_to = "Sample", values_to = "Log2CPM")
ggplot(dfLogLongerp1, aes(x = Log2CPM, color=Sample)) +
  geom_density() +
  labs(x = "Log2CPM", y = "Density") +
  theme_minimal() + 
  ggtitle("Density Plot showing Log2 CPM Densities Prenormalisation")
```

## Normalisation 
```{r message = FALSE, warning = FALSE}
# Plot for the before and after
groups <- c("Mock", "Mock", "Mock", "Control", "Control", "Control", "GW-Treated", "GW-Treated", "GW-Treated")

ofDGE <- edgeR::DGEList(counts = cleanData, group=groups, remove.zeros = TRUE)
ofDGE <- edgeR::calcNormFactors(ofDGE)

normalised <- edgeR::cpm(ofDGE)
saveRDS(normalised, "NormalisedData.RDS")
write.csv(normalised, "normalised_data.csv")


## BOXPLOT
### Adding 1 to prevent 0 expression from going to -inf
dataAdd1n <- normalised + 1
dataLogn <- log2(edgeR::cpm(dataAdd1n))

dfLogn <- as.data.frame(dataLogn)
dfLogn$Gene <- rownames(dfLogn)
dfLogLongern <- tidyr::pivot_longer(dfLogn, cols = -Gene, names_to = "Sample", values_to = "Log2CPM")

ggplot2::ggplot(dfLogLongern, aes(x = Sample, y = Log2CPM)) +
  geom_boxplot() + 
  labs(x = "Samples", y = "Log2 CPM") +
  ggtitle("Boxplot showing Log2 CPM Post-Normalisation")
  
## MDS PLOT
limma::plotMDS(log2(normalised), labels=NULL, pch=1,
               col = c("red", "green", "blue")[factor(groups)])

legend("topright",
  legend=levels(factor(groups)),
  pch=c(1), col= c("red", "green","blue"),title="Class",
  bty = 'n', cex = 0.75)

## DENSITY PLOT
## Add 0.1 to each value in the normalised data to prevent 0 expression error
dataAddp1n <- normalised + 0.1
dataLogp1n <- log2(edgeR::cpm(dataAddp1n))

dfLogp1n <- as.data.frame(dataLogp1n)
dfLogp1n$Gene <- rownames(dfLogp1n)
dfLogLongerp1n <- tidyr::pivot_longer(dfLogp1n, cols = -Gene, names_to = "Sample", values_to = "Log2CPM")
ggplot(dfLogLongerp1, aes(x = Log2CPM, color=Sample)) +
  geom_density() +
  labs(x = "Log2CPM", y = "Density") +
  theme_minimal() + 
  ggtitle("Density Plot showing Log2 CPM Densities Post-Normalisation")
```

## BIBLIOGRAPHY
